---
title: "Working with TRIM command files and TRIM data files"
author: "Patrick Bogaart, Mark van der Loo, and Jeroen Pannekoek"
date: "23 November 2017"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Working with TRIM command files and TRIM data files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## TRIM Command Files

The original TRIM software can be controlled with text files containing a series of commands that specify both the location and format of the data, an the model (or models) to compute. Such TRIM command files (usually stored with the extension `.tcf`) should be considered legacy but for backwards compatability they can be used from R.

To try this, execute the code below to create a `tcf` file and a TRIM data file in the current
working directory of R.
```{r}
library(rtrim)
tmp <- "FILE skylark.dat
TITLE  skylark-1d
NTIMES 8
NCOVARS 2
LABELS
Habitat
Cov2
END
MISSING 999
WEIGHT Absent
COMMENT Example 1; using linear trend model
WEIGHTING off
OVERDISP on
SERIALCOR on
MODEL 2
"
write(tmp,file="skylark.tcf")
data(skylark)
skylark[is.na(skylark)] <- 999
write.table(skylark,file="skylark.dat",col.names=FALSE,row.names=FALSE)
```

Executing a TRIM command file is as easy as reading the file using `read_tcf` and passing the result to `trim`.

```{r}
tc <- read_tcf("skylark.tcf")
m <- trim(tc)
```
The resulting `trim` object can be evaluated as described in the [Getting Started](rtrim_for_TRIM_users.html) vignette. For example
```{r}
wald(m)
``` 


The object `tc`, resulting from `read_tcf` is an object of class `trimcommand`. It stores all commands defined in the TRIM command file. Note that logical parameters such as `WEIGHT` are transformed to `logical` in R.
```{r}
tc
```


**NOTE.** Be aware that R has its own present working directory. If relative paths (that is, file names not starting with the full path to their location) are used in the TRIM command file, R will interpret them as relative to the current working directory.

## TRIM data files

TRIM data files are basically space-separated, tabular texstfiles where the order and type of columns is fixed by a few parameters. Given such a specification, a file can be read with `read_tdf`. 
